Data Analysis Methods for the Detection of the Epoch of Reionization

### 2017-07-18 PAPER.tex
-Fleshed out Section 5.2 (data bootstrapping)

In this section, we highlight two major changes since \citet{ali_et_al2015}, using the lessons we've learned about sampling with replacement and bootstrapping independent samples.
\cc{Include info about baseline bootstrapping:}
For the \citet{ali_et_al2015} method, each group is then sampled with replacement to create a new group, of the same size, that can have repeated baselines inside it. We discover that in doing so, we are sacrificing some of our sensitivity since this results in there being 3-4 repeated baselines per group. In order to maximize our sensitivity but still apply random sampling for use in error estimation, we instead form new groups using all independent baselines except the very last one. For example, if we have $10$ baselines in a group, we use the first $9$ to guarantee at least $9$ independent measurements, and then fill the last slot randomly out of the $10$. We do this for all $5$ groups. This is still a valid means of bootstrapping because there are many more possibilities of baseline groupings than the number of bootstraps we run for this analysis ($nboots = 20$).

\cc{Include info about second round of bootstrapping:}
In \citet{ali_et_al2015}, a second round of bootstrapping occurs over the bootstrap and time axes simultaneously. Random values are sampled with replacement along both axes, drawing as many values as there are number of bootstraps and times. Final power spectrum limits are then computed by taking the mean and standard deviation over this second bootstrap axis.

However, we have found that this method greatly underestimates power spectrum errors, especially for fringe-rate filtered data. This can be explained by the fact that fringe-rate filtered data has a dramatically reduced number of independent modes. Hence, drawing $100$ samples out of a length-$100$ dataset that only has $5$ independent modes in it, for example, results in a narrower distribution of values that leads to a false error estimation.

To avoid this issue, we instead take a simple average along the time axis. Our final power spectrum limits are computed by taking the mean and standard deviation over our single bootstrap axis.

>>>>

In this section, we highlight two major changes since \citet{ali_et_al2015}, using the lessons we've learned about sampling with replacement and bootstrapping independent samples.

As discussed in Section \ref{sec:ErrorOverview}, traditional bootstrapping theory is only a valid way of estimating errors if a dataset is comprised of independent samples. Recall that the PAPER-64 pipeline outputs $20$ bootstraps (over baselines), each a $2$-dimensional power spectrum that is a function of $k$ and time. In \citet{ali_et_al2015}, a second round of bootstrapping occurs over the baseline bootstrap and time axes simultaneously. Random values are sampled with replacement along both axes, drawing as many values as there are number of bootstraps and times. Final power spectrum limits are then computed by taking the mean and standard deviation over this second bootstrap axis (which is typically large, on the order of $N_{boot} \sim 400$).

Because our dataset is fringe-rate filtered, we have substantially increased our integration time, resulting in fewer independent samples (from $\sim700$ samples to $8$ samples). A random draw of $700$ samples from this dataset therefore has many repeated samples, and the variance between hundreds ($N_{boot}$) of these random samples is smaller than the true underlying variance. Another way of describing this underestimation is that there are fewer ways to randomly sample data with $8$ independent modes versus $700$, so bootstrapping a fringe-rate filtered dataset yields a narrower distribution of values from which we calculate its standard deviation.

To prevent error underestimation, we simply take the average along the time axis in our revised pipeline rather than bootstrapping. Hence, our final power spectrum errors are calculated from the standard deviation over the baseline bootstrapping axis only, which is still a valid axis since each baseline makes an independent measurement of the sky.

The second aspect of bootstrapping discussed in Section \ref{sec:ErrorOverview} is in regard to how a dataset is randomly sampled. In order to maximize sensitivity (i.e. maximize independent samples in a bootstrap), we have revised the method by which we sample baselines when bootstrapping.

First, we speed things up in our pipeline by dividing our total number of baselines into $5$ groups, where there are no repeated baselines within or between groups. In \citet{ali_et_al2015}, each group is then sampled with replacement to create a new group of the same size, which can have repeated baselines inside it. We discover that in doing so, we are sacrificing some of our sensitivity since there are $3$-$4$ repeated baselines per group. In order to maximize our sensitivity but still apply random sampling for use in error estimation, we instead form new groups using all independent baselines except the very last one, which we fill randomly. We use this method for every bootstrap, creating new groups and sampling the last spot randomly each time.

Power spectrum estimates for PAPER-64 fringe-rate filtered data are shown in Figure \ref{fig:data_errors}. The estimates use a weighting matrix of $\textbf{C}_{eff}^{-1}$, where $\textbf{C}_{eff} = \textbf{C}*\textbf{I}$, in order to minimize signal loss. We show upper limits ($2\sigma$ error bars) for three different results. Our revised error estimation method is in black, where we only bootstrap along the baseline axis and we maximize sensitivity by random sampling only the last baseline in each group. The red curve uses a similar method except all baselines are randomly sampled, leading to a higher power spectrum estimate because some baselines are repeated per group. In blue, we bootstrap along both the baseline and time axes, drawing as many samples as there are in the data. We under-estimate errors in doing so because we are drawing many more samples than independent ones since the data is fringe-rate filtered.

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{plots/data_errors.png}
	\caption{$2\sigma$ upper limit power spectrum estimates using PAPER-64 data and three different bootstrapping methods. The data is fringe-rate filtered and a weighting matrix of $\textbf{C}_{eff}^{-1}$ is used, where $\textbf{C}_{eff} = \textbf{C}*\textbf{I}$, in order to minimize signal loss. Three different bootstrapping methods are shown: the black and red use methods where bootstrapping occurs on the baseline axis only (not the time axis), and they differ by how the baselines are randomly sampled. It is evident that sensitivity is maximized when baselines are sampled in a way that ensures mostly independent samples (only the last baseline spot is sampled randomly). The blue result bootstraps over both the baseline and time axes, and illustrates how errors can be under-estimated if sampling more values than independent ones (fringe-rate filtering reduces the number of independent samples).}
	\label{fig:data_errors}
\end{figure}
